Hydra battles in Coq

### 2020-09-23 doc/hydras.tex
minor modifications

\include{chapter-introduction}
\include{chapter-hydras}
\include{chapter-ordinals}
\include{chapter-termination}
\include{chapter-ks}
\include{chapter-large-sets}
\include{chapter-schutte}

>>>>

%-------------------------------------------------------------------

\chapter{Introduction}

\vspace{16pt}

Proof assistants are excellent tools for exploring the structure of mathematical proofs,
studying  which hypotheses are really needed, and which proof patterns are useful and/or
necessary. Since the development of a theory is represented as a bunch of computer files,
everyone is able to read the proofs with an arbitrary level of detail, or to play with the theory by building alternate proofs or definitions.

Among all the theorems proved with the help of proof assistants like \coq{}, \isabelle{}, \hol{}, etc.,
several statements and proofs  share some interesting features:
\begin{itemize}
\item Their statements are easy to understand, even by non-mathematicians
\item Their proof requires some non-trivial mathematical tools
\item Their mechanization on computer presents some methodological interest.
\end{itemize}

This is obviously the case of the four-color theorem~\cite{fourcolors}  and the Kepler conjecture~\cite{flyspeck2015}. We do not mention impressive works like the proof of the odd-order theorem ~\cite{oddorderthm}, since understanding its statement requires a quite good mathematical culture.

Hydra games (a.k.a. \emph{Hydra battles}) appeared in an article published in 1982 by two mathematicians:
L. Kirby and J. Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}. 
Although the mathematical contents of this 
paper are quite advanced, the rules of hydra battles are very easy to understand. There are now several sites on Internet where you can find tutorials on hydra games, together with simulators you can play with. See, for instance, the page written by Andrej Bauer~\cite{bauer2008}.

Hydra battles, as well as Goodstein Sequences~\cite{goodstein_1944, KP82}
are a nice way to present complex termination problems.
The article by Kirby and Paris presents a proof of termination
based on ordinal numbers, as well as a proof that this termination is not
provable in Peano arithmetic. In the book dedicated to 
J.P. ~Jouannaud \cite{HommageJPJ}, N.~Dershowitz and G.~Moser  give a thorough survey on this topic~\cite{Dershowitz2007}.

Here, we present a development  for the \coq{} proof assistant, after the work of Kirby and Paris. This formalization contains the following main parts:

\begin{itemize}
\item Representation in \coq{} of hydras and hydra battles
\item A proof that every battle is finite and won by Hercules. This proof is based on a \emph{variant} which maps any hydra to an ordinal strictly less than $\epsilon_0$ and is strictly decreasing along any battle.

\item Using a combinatorial toolkit designed by J.~Ketonen and R.~Solovay~\cite{KS81}, we prove that, for any ordinal $\mu<\epsilon_0$, there exists no such variant mapping any hydra to an ordinal stricly less than $\mu$. Thus, the complexity of $\epsilon_0$ is really needed in the previous proof.

\item We prove a relation between the length of a ``classic''  class of  battles
and the Wainer-Hardy hierarchy of ``rapidly growing functions'' $H_\alpha$~\cite{Wainer1970}. The considered class of battles, which we call \emph{standard}  is the most considered one in the scientific  litterature(including popularization).
\end{itemize}

Simply put, this document tries to combines the scientific interest of two articles~\cite{KP82, KS81} and a book~\cite{schutte} with the playful activity of truly proving theorems.
We hope  that such a work, besides exploring a nice piece of discrete maths, 
will show how \coq{} and its standard library are well fitted to help us to understand some non-trivial mathematical developments, and also to experiment the constructive parts of  the proof through functional programming.

\begin{figure}[h]
  \centering
  \includegraphics[width=11cm]{epsilon0.jpg}
  \caption{Discrete mathematics and formal proofs}
  \label{fig:gamma0}
\end{figure}
 We also hope to provide a little clarification on infinity (both potential and actual) through the notions of function, computation, limit,
 types and proofs.

\section{Remarks}

\subsection{Difference from Kirby and Paris's work}
In~\cite{KP82}, Kirby and Paris showed  that there is no proof of termination of all hydra battles in Peano Arithmetic (PA).
Since we are used to writing proofs in higher order logic, the restriction to PA was quite unnatural for us. So we chosed to prove another statement without any reference to PA, by considering a class of proofs indexed by ordinal numbers upto $\epsilon_0$.

\subsection{Trust in our proofs?}
\label{sect:trust-in-proofs}

Unlike mathematical literature, where definitions and proofs are spread over many articles and books,
the whole proof is now inside your computer. It is composed of the \texttt{.v} files you downloaded and 
parts of \coq's standard library. Thus, there is no ambiguity in our definitions and the premises of the theorems. Furthermore, you will be able to navigate through the development, using your favourite text editor or IDE, and some commands like \texttt{Search}, \texttt{Locate},  etc.

\subsection{Assumed redundancy}

It may often happen that several definitions of a given concept, or several proofs of a given theorem are possible. If all the versions present some interest, we will make them available, since each one may be of some methodological 
interest (by illustrating some tactic of proof pattern, for instance).
We may use \coq's module system to make several proofs of a givent theorem co-exists in our libraries (see also Sect~\vref{sect:alt-proofs}).
After some discussions of the pros and cons of each solution, we will choose to develop only one of them, leaving the others  as exercises or projects (i.e., big or difficult exercises).
In order to discuss which assumptions are really needed for proving a theorem, we will also present 
several aborted proofs.
Of course, do not hesitate to contribute nice proofs or alternative definitions !

It may also happen that some proof looks to be useless, because the proven theorem is a trivial consequence of another (proven too) result.
For instance, let us consider the three following statements:
\begin{enumerate}
\item There is no measure into $\mathbb{N}$ for proving the termination of all hydra battles (Sect~\vref{omega-case}).
\item There is no measure into $[0,\omega^2)$ for proving the termination of all hydra battle (Sect~\vref{omega2-case}).
\item There is no measure into $[0,\mu)$ for proving the termination of all hydra battles, for any $\mu<\epsilon_0$ (Sect\vref{sec:free-battles-case}).
\end{enumerate}

Obviously, the third theorem implies the second one, which implies the first one. So, theoretically, a library would contain only a proof of $(3)$ and remarks for $(2)$ and $(1)$. But we found it interesting to make all the three proofs available, allowing the reader to compare their common structure and notice their technical differences.
In particular, the proof of $(3)$ uses several non-trivial combinatorial properties of ordinal numbers~\cite{KS81}, whilst $(1)$ and $(2)$ use simple properties of $\mathbb{N}$ and $\mathbb{N}^2$.

\subsection{About Logic}

Most of the proofs we present are \emph{constructive}. Whenever possible, we provide the user with an associated function, which she or he can apply in \gallina{} or \ocaml{} in order to get a ``concrete'' feeling of the meaning of the considered theorem.
For instance, in Chapter~\vref{chap:ketonen}, the notion of \emph{limit ordinal} is
made more ``concrete'' thanks to a function \texttt{canon} which computes every item of a sequence which converges on a given limit ordinal $\alpha$. This simply typed function allows the user/reader to make her/his own experimentations.
For instance, one can very easily compute the $42$-nd item of a sequence which converges towards $\omega^{\omega^\omega}$.

Except in the \texttt{Schutte} library, dedicated to an axiomatic presentation of the set of countable ordinal numbers, all our development is axiom-free, and respects the rules of intuitionistic logic. Note that we also use the \texttt{Equations} plug-in~\cite{sozeau:hal-01671777} in the definitition of  several rapidly growing hierarchy of functions, in Chap.~\ref{chap:alpha-large}. This plug-in imports several known-as-harmless  axioms.

% \begin{Coqsrc}
% FunctionalExtensionality.functional_extensionality_dep : 
% forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
% (forall x : A, f x = g x) -> f = g
% \end{Coqsrc}

\index{Coq!Commands!Print Assumptions}

At any place of our development, you may use the command \texttt{Print Assumptions {\it ident}} in order to verify on which axiom the theorem {\it ident} may depend. The following example is extracted from 
Library~\href{../src/html/hydras.Epsilon0.F_alpha.html}{hydras.Epsilon0.F\_alpha}, where we use the \texttt{coq-equations} plug-in (see Sect.~\vref{sect:wainer}).

\begin{Coqsrc}
About F_zero_eqn.
\end{Coqsrc}

\begin{Coqanswer}
F_zero_eqn : forall i : nat, F_ Zero i = S i
\end{Coqanswer}

\begin{Coqsrc}
Print Assumptions F_zero_eqn. 
\end{Coqsrc}

\begin{Coqanswer}
Axioms:
FunctionalExtensionality.functional_extensionality_dep
  : forall (A : Type) (B : A -> Type) (f g : forall x : A, B x),
    (forall x : A, f x = g x) -> f = g
Eqdep.Eq_rect_eq.eq_rect_eq
  : forall (U : Type) (p : U) (Q : U -> Type) (x : Q p) (h : p = p),
    x = eq_rect p Q x p h
\end{Coqanswer}

\subsection{Main references}

In our development, we adapt the definitions and prove many theorems which
we found in the following articles. 
\begin{itemize}
\item ``Accessible independence results for Peano arithmetic''  by Laurie Kirby and Jeff Paris~\cite{KP82}
\item ''Rapidly growing Ramsey Functions'' by Jussi Ketonen and Robert Solovay~\cite{KS81}
\item ``The Termite and the Tower'', by Will Sladek~\cite{Sladek07thetermite}
\item Chapter V of ``Proof Theory'' by Kurt Schütte~\cite{schutte}
\end{itemize}

\paragraph*{Warning:}

This document is \emph{not} an introductory text for \coq{}, and there are many aspects of this proof assistant that are not covered. 
 The reader should already have some basic experience with the \coq{} system. The Reference Manual and several tutorials are available on \coq{} page~\cite{Coq}.  First chapters of textbooks like \emph{Interactive Theorem Proving and Program Development}~\cite{BC04}, \emph{Software Foundations}~\cite{SF} or  \emph{Certified Programming with Dependent Types} ~\cite{chlipalacpdt2011} will give you the right background.

\subsection{State of the development}
The \coq{} scripts herein are in constant development since our contribution~\cite{CantorContrib} on ordinal notations for the ordinals $\epsilon_0$ and $\Gamma_0$.
We added new material : axiomatic definition of countable ordinals after Schütte~\cite{schutte}, combinatorial aspects of $\epsilon_0$, after Ketonen and Solovay~\cite{KS81} and Kirby and Paris~\cite{KP82}, recent \coq{} technology: type classes, equations, etc.

We are now working in order to make clumsy proofs more readable, simplify definitions, 
to ``factorize'' proofs as much as possible.

Many ``todo''s or projects are put in this text, suggesting possible improvements.

\subsection{Contributions}
Many thanks to \'Evelyne Contejean and Florian Hatat for their contribution. \'Evelyne contributed libraries on the recursive path ordering (\emph{rpo}) for proving the well-foundedness of our representation of $\epsilon_0$ and $\Gamma_0$.
Florian Hatat proved many useful lemmas on countable sets, which we used in our adaptation of Schütte's formalization of countable ordinals.

\label{sec:orgheadline2}

Any form of contribution  is welcome: correction of errors, improvement of
Coq scripts, proposition of inclusion of new chapters, and generally any
comment or proposition that would help us. The text contains several \emph{projects} which, when completed, may improve the present work.
Please do not hesitate to bring your contribution!

\subsection{Typographical conventions}

Quotations from our  \coq{} source are displayed as follows:

  \begin{Coqsrc}
 Require Import Arith.

 Definition square (n:nat) := n * n.

 Lemma square_double : exists n:nat, n + n = square n.
 Proof.
    exists 2. 
  \end{Coqsrc}

Answers from \coq{} (including sub-goals, error messages, etc.) are displayed in slanted style
with a different background color.

 \begin{Coqanswer}
 1 subgoal, subgoal 1 (ID 5)

  ============================
   2 + 2 = square 2

 \end{Coqanswer}

 \begin{Coqsrc}
   reflexivity.
Qed.
 \end{Coqsrc}

\subsection{Active links}
The  links which appear in this pdf  document lead are of three possible kinds of destination:
\begin{itemize}
\item Local links to the document itself,
\item External links, mainly to \coq's page,
\item Local links to pages generated by \texttt{coqdoc}. According to the current makefile,
  we assume that the page generated from a library \texttt{XXX/YYY.v} is stored as
the relative address \texttt{../src/html/hydras.XXX.YYY.html} (from the location of the pdf).
Thus,  active links towards our \coq{} modules may be incorrect if you got this \texttt{pdf} document otherwise than by compiling the distribution available in
\url{https://framagit.org/casteran/hydra-battles-in-coq}.

\end{itemize}

\subsection{Alternative or bad definitions}
\label{sect:alt-proofs}
Finally, we decided to include definitions or lemma statements, as well as tactics,  that lead to
dead-ends or to too complex developments, with the following coloring.
Bad definitions 
 and encapsulation in modules called \texttt{Bad}, \texttt{Bad1}, etc.

\begin{Coqbad}
Module Bad.

Definition double (n:nat)  := n + 2.

Lemma lt_double : forall n:nat, n < double  n.
Proof.
   unfold double; omega.
Qed.

End Bad.
\end{Coqbad}

Likewise, alternative, but still unexplored definitions will be presented in modules
\texttt{Alt}, \texttt{Alt1}, etc. Using these definitions is left as an implicit exercise.

\begin{Coqalt}
Module Alt.

  Definition double (n : nat) := 2 * n.

End Alt.
\end{Coqalt}

\begin{Coqsrc}
Lemma alt_double_ok n : Nat.double n = Alt.double n.
Proof.
  unfold Alt.double, Nat.double.
  omega.
Qed.
\end{Coqsrc}

\section{Acknowledgements}
\label{sec:orgheadline5}
    Many thanks to David Ilcinkas, Sylvain Salvati, Alan Schmitt and Théo Zimmerman for their help on the elaboration of this document, and to the
 members of the \emph{Formal Methods} team at laBRI for their helpful comments 
on an oral presentation of this work.

Many thanks also to the Coq development team, Yves Bertot, and members of the \emph{Coq Club} for interesting discussions about the \coq{} system and the Calculus of Inductive Constructions.

The author of the present document wishes to express his gratitude to the late Patrick Dehornoy, whose talk  was determinant for our desire to work on this topic.

I owe my interest in discrete mathematics and their relation to formal proofs and functional programming  to Srecko Brlek.  Equally, there is W. H. Burge's book ``\emph{Recursive Programming Techniques}'' ~\cite{burge} which was a great  source of inspiration.

% \subsubsection{Links to the Coq source}

%-----------------------------------------------------------

\chapter{Hydras and Hydra Games}
\label{sec:orgheadline91}

This chapter is dedicated to the representation of hydras and rules of the hydra game in \coq's specification language:\gallina.

Technically, a \emph{hydra} is just a finite ordered tree, each node of which 
has any number of sons. Note that, contrary to the computer science tradition, we will show the hydras 
with the heads up and the foot (i.e., the root of the tree) down.
Fig.~\ref{fig:Hy} represents such  a hydra, which will be referred to as \texttt{Hy} in our examples (please look at the 
module~\href{../src/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {$\bullet$};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H2) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H4) at (4,2) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N3) to [bend left= 10]  (H1) ;
\draw (N3) to [bend right= 16] (H2);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (foot) to [bend right = 10] (H4) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{The hydra Hy \label{fig:Hy}}
\end{figure}

We use a specific vocabulary for talking about hydras. Table~\ref{tab:hyd2tree} shows the correspondance between our terminology and the usual vocabulary for trees in computer science.

\begin{figure}[h]
  \centering
  \begin{tabular}{ll}
Hydras & Finite rooted trees\\
\hline
foot & root\\
head & leaf\\
node & node\\
segment  & (directed) edge \\
sub-hydra & subtree\\
daughter & immediate subtree\\
\end{tabular}
  \caption{Translation from hydras to trees}
  \label{tab:hyd2tree}
\end{figure}

The hydra \texttt{Hy} has a \emph{foot} (below), five \emph{heads}, and eight \emph{segments}. 
We leave it to the reader to define various parameters such as the height, the size, the highest arity (number of sons of a node) of a hydra. In our example, these parameters have the respective values : $4$, $9$ and $3$.

\subsection{The rules of the game}

\label{sec:orgheadline44}
A \emph{hydra battle} is a fight between Hercules and the Hydra. 
More formally, a  battle is a sequence of \emph{rounds}.
At each round:
\begin{itemize}
\item If the hydra is composed of just one head, the battle is finished
and  Hercules is the winner.
\item Otherwise, Hercules chops off \emph{one} head of the hydra,

\begin{itemize}
\item If the head is at distance 1 from the foot, the head is just lost by the hydra, with no more reaction.
\item Otherwise, let us denote by \(r\) the node that was at distance \(2\) from 
the removed head in the direction of the foot,  and consider the  sub-hydra \(h'\) of \(h\), whose  root is \(r\) \footnote{$h'$ will be called ``the wounded part of the hydra'' in the subsequent text. In Figures~\vref{fig:Hy2} and ~\vref{fig:Hy4}, this sub-hydra  is displayed in red.}. Let $n$ be some natural number.
Then $h'$ is replaced by  $n+1$ of copies of \(h'\) which share the same root $r$.
 The \emph{replication number} $n$ may be different (and generally is)   at each round of the fight.
It may be chosen by the hydra, according to its strategy, or imposed by some 
particular rule. In many presentations of hydra battles, this number is increased by $1$ at each round. In the following presentation, we will also consider battles where the hydra is free to chose its number of replication at every round of the battle\footnote{Let us recall that, if the chopped-off head was at distance 1 from the foot, the replication factor is meaningless.}.
\end{itemize}
\end{itemize}

Note that the description given in~\cite{KP82} of the replication process in hydra battles is also  semi-formal.

\label{original-rules}

\begin{quote}
  ``From the node that used to be attached to the head which was just chopped off, traverse one 
segment towards the root until the next node is reached. From this node sprout $n$ replicas of 
that part of the hydra (after decapitation) which is ``above'' the segment just traversed, i.e., those 
nodes and segments from which, in order to reach the root, this segment would have to be 
traversed. If the head just chopped off had the root of its nodes, no new head is grown. ''
\end{quote}

Moreover, we note that this description is in \emph{imperative} terms. In order to build a formal  study of the properties of hydra battles, we prefer to use a mathematical vocabulary, i.e., graphs, relations, functions, etc.
Thus, the replication process will be represented as a binary relation on a data type \texttt{Hydra},
linking the state of the hydra \emph{before} and \emph{after} the transformation.
A battle will thus be represented as a sequence of terms of type \texttt{Hydra}, respecting the rules of the game.

\subsection{Example}
Let us start a battle between Hercules and the hydra \texttt{Hy} of Fig.~\ref{fig:Hy}.

At the first round, Hercules choses to chop off the rightmost head of \texttt{Hy}.
Since this head is near the floor, the hydra loses this head. Let us call 
 \texttt{Hy'} the resulting state of the hydra, represented in Fig.~\vref{fig:Hy-prime}.

Next, assume Hercules choses to chop off one of the two highest heads of \texttt{Hy'}, for instance the rightmost one. Fig.~\vref{fig:Hy2} represents the rotten neck in dashed lines, and the part that will be replicated in red. Assume also that the hydra decides to add 4 copies of the red part\footnote{In other words, the replication factor at this round is equal to $4$.}. We obtain a new state \texttt{Hy''} depicted in Fig.~\ref{fig:Hy3}.

\begin{figure}[h]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {$\bullet$};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H2) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\node (H4) at (6,0) {$\Xey[2][lightgray]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3) ;
\draw (N3) to [bend left= 10]  (H1) ;
\draw (N3) to [bend right= 16] (H2);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}

\caption{Hy': the state  of Hy after one round \label{fig:Hy-prime}}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.5]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4)  {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
\node (H2) at (5,0) {$\Xey[2][lightgray]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\node (ex) at (5,8) {};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw  (N2) -- (N3)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw [dashed] (N3) to [bend left= 10](ex);
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend right= 16] (H5);
\end{tikzpicture}
\caption{A second beheading}
\label{fig:Hy2}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0) {$\bullet$};
\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H1) at (0,8) {$\Smiley[2][vertfluo]$};
\node (H11) at (2,8) {$\Smiley[2][vertfluo]$};
\node (H12) at (4,8) {$\Smiley[2][vertfluo]$};
\node (H13) at (6,8) {$\Smiley[2][vertfluo]$};
\node (H14) at (8,8) {$\Smiley[2][vertfluo]$};

\node (N3) at (1,6) {$\bullet$};
\node (N31) at (2,6) {$\bullet$};
\node (N32) at (3,6) {$\bullet$};
\node (N33) at (4,6) {$\bullet$};
\node (N34) at (5,6) {$\bullet$};

\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,4) {$\Smiley[2][vertfluo]$};
\draw (foot) -- (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N2) -- (N31);
\draw (N2) -- (N32);
\draw (N2) -- (N33);
\draw (N2) -- (N34);
\draw (N3) to   [bend left= 10](H1) ;
\draw (N31) to   [bend left= 10](H11) ;
\draw (N32) to   [bend left= 10](H12) ;
\draw (N33) to   [bend left= 10](H13) ;
\draw (N34) to   [bend left= 10](H14) ;
\draw (foot) to [bend left= 10]  (H0) ;
\draw (N1) to [bend left= 10]  (H5) ;
\end{tikzpicture}
\caption{Hy'', the state of Hy after two rounds \label{fig:Hy3}}
\end{figure}

Figs.~\ref{fig:Hy4} and~\vref{fig:Hy5} represent a possible third round of the battle, with a replication factor equal to $2$. Let us call \texttt{Hy'''} the state of the hydra after that third round.

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.6]

\node (foot) at (2,0)  {{\color{lightred}$\bullet$}};
\node (N1) at (2,2) {{\color{lightred}$\bullet$}};
\node (N2) at (2,4) {{\color{lightred}$\bullet$}};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (exN4) at (4,4) {};
\node (H1) at (0,8) {$\Sey[2][lightred]$};
\node (H11) at (2,8) {$\Sey[2][lightred]$};
\node (H12) at (4,8) {$\Sey[2][lightred]$};
\node (H13) at (6,8) {$\Sey[2][lightred]$};
\node (H14) at (8,8) {$\Sey[2][lightred]$};

\node (N3) at (1,6) {{\color{lightred}$\bullet$}};
\node (N31) at (2,6) {{\color{lightred}$\bullet$}};
\node (N32) at (3,6) {{\color{lightred}$\bullet$}};
\node (N33) at (4,6) {{\color{lightred}$\bullet$}};
\node (N34) at (5,6) {{\color{lightred}$\bullet$}};

\node (H0) at (0,2) {$\Smiley[2][vertfluo]$};
\node (H5) at (4,0) {$\Xey[2][lightgray]$};
\draw (foot) -- (N1)[very thick,draw=lightred] ;
\draw (N1) -- (N2)[draw=lightred];
\draw (N2) -- (N3)[draw=lightred];
\draw (N2) -- (N31)[draw=lightred];
\draw (N2) -- (N32)[draw=lightred];
\draw (N2) -- (N33)[draw=lightred];
\draw (N2) -- (N34)[draw=lightred];
\draw (N3) to   [bend left= 10](H1) [draw=lightred];
\draw (N31) to   [bend left= 10](H11) [draw=lightred];
\draw (N32) to   [bend left= 10](H12) [draw=lightred];
\draw (N33) to   [bend left= 10](H13) [draw=lightred];
\draw (N34) to   [bend left= 10](H14) [draw=lightred];
\draw (foot) to [bend left= 10]  (H0) ;
\draw [dashed] (N1) to  [bend left= 10](exN4);
\end{tikzpicture}
\caption{A third beheading (wounded part in red) \label{fig:Hy4}}
\end{figure}

\begin{figure}[hp]
\centering
\begin{tikzpicture}[very thick, scale=0.4]

\node (foot) at (10,0) {$\bullet$};

\node (N1) at (2,2) {$\bullet$};
\node (N2) at (2,4) {$\bullet$};
\node (N3) at (2,6) {{\color{lightred}$\bullet$}};
\node (H1) at (0,8) {$\Smiley[1][vertfluo]$};
\node (H11) at (2,8) {$\Smiley[1][vertfluo]$};
\node (H12) at (4,8) {$\Smiley[1][vertfluo]$};
\node (H13) at (6,8) {$\Smiley[1][vertfluo]$};
\node (H14) at (8,8) {$\Smiley[1][vertfluo]$};

\node (N3) at (1,6) {$\bullet$};
\node (N31) at (2,6) {$\bullet$};
\node (N32) at (3,6) {$\bullet$};
\node (N33) at (4,6) {$\bullet$};
\node (N34) at (5,6) {$\bullet$};

\node (H0) at (-3,3) {$\Smiley[1][vertfluo]$};

\draw (foot) to [bend left=10] (N1)[very thick] ;
\draw (N1) -- (N2);
\draw (N2) -- (N3);
\draw (N2) -- (N31);
\draw (N2) -- (N32);
\draw (N2) -- (N33);
\draw (N2) -- (N34);
\draw (N3) to   [bend left= 10](H1) ;
\draw (N31) to   [bend left= 10](H11) ;
\draw (N32) to   [bend left= 10](H12) ;
\draw (N33) to   [bend left= 10](H13) ;
\draw (N34) to   [bend left= 10](H14) ;
\draw (foot) to [bend left = 15]  (H0) ;

% second copy 
\node (N01) at (12,2) {$\bullet$};
\node (N02) at (12,4) {$\bullet$};
\node (N03) at (12,6) {{\color{lightred}$\bullet$}};
\node (H001) at (10,8) {$\Smiley[1][vertfluo]$};
\node (H0011) at (12,8) {$\Smiley[1][vertfluo]$};
\node (H0012) at (14,8) {$\Smiley[1][vertfluo]$};
\node (H0013) at (16,8) {$\Smiley[1][vertfluo]$};
\node (H0014) at (18,8) {$\Smiley[1][vertfluo]$};

\node (N03) at (11,6) {$\bullet$};
\node (N031) at (12,6) {$\bullet$};
\node (N032) at (13,6) {$\bullet$};
\node (N033) at (14,6) {$\bullet$};
\node (N034) at (15,6) {$\bullet$};

\draw (foot) -- (N01)[very thick] ;
\draw (N01) -- (N02);
\draw (N02) -- (N03);
\draw (N02) -- (N031);
\draw (N02) -- (N032);
\draw (N02) -- (N033);
\draw (N02) -- (N034);
\draw (N03) to   [bend left= 10](H001) ;
\draw (N031) to   [bend left= 10](H0011) ;
\draw (N032) to   [bend left= 10](H0012) ;
\draw (N033) to   [bend left= 10](H0013) ;
\draw (N034) to   [bend left= 10](H0014) ;

% third copy 
\node (N001) at (22,2) {$\bullet$};
\node (N002) at (22,4) {$\bullet$};
\node (N003) at (22,6) {{\color{lightred}$\bullet$}};
\node (H001) at (20,8) {$\Smiley[1][vertfluo]$};
\node (H0011) at (22,8) {$\Smiley[1][vertfluo]$};
\node (H0012) at (24,8) {$\Smiley[1][vertfluo]$};
\node (H0013) at (26,8) {$\Smiley[1][vertfluo]$};
\node (H0014) at (28,8) {$\Smiley[1][vertfluo]$};

\node (N003) at (21,6) {$\bullet$};
\node (N0031) at (22,6) {$\bullet$};
\node (N0032) at (23,6) {$\bullet$};
\node (N0033) at (24,6) {$\bullet$};
\node (N0034) at (25,6) {$\bullet$};

\draw (foot) -- (N001)[very thick] ;
\draw (N001) -- (N002);
\draw (N002) -- (N003);
\draw (N002) -- (N0031);
\draw (N002) -- (N0032);
\draw (N002) -- (N0033);
\draw (N002) -- (N0034);
\draw (N003) to   [bend left= 10](H001) ;
\draw (N0031) to   [bend left= 10](H0011) ;
\draw (N0032) to   [bend left= 10](H0012) ;
\draw (N0033) to   [bend left= 10](H0013) ;
\draw (N0034) to   [bend left= 10](H0014) ;
\end{tikzpicture}
\caption{The configuration Hy''' of Hy \label{fig:Hy5}}
\end{figure}
\FloatBarrier

We leave it to the reader  to guess the following  rounds of the battle \dots
 % Please keep in mind that, in this 
% the hydra is free to chose any number of replications at each time, whereas
% Hercules chops only one head per round.

% Let us precise that, in this game, Hercules wins if the hydra is eventually reduced 
% to a single head. 
% We know from~\cite{KP82} that, whichever the initial configuration of the
% hydra, and the strategies of both players, Hercules eventually wins. The 
% aforementionned paper shows also that there do not exist any \emph{simple} proof of this result.

\section{Hydras and their representation in \emph{Coq}}
\label{sec:orgheadline48}

In order to describe trees where each node can have an arbitrary (but finite) number of sons, it usual to define a type where each node carries a \emph{forest}, \emph{i.e} a list of trees
(see for instance Chapter 14, pages 400-406 of \cite{BC04}).

For this purpose, we define two mutual \emph{ad-hoc}  inductive types, where \texttt{Hydra} is the main type, and \texttt{Hydrae} is a helper for describing finite sequences of hydra.
\label{types:Hydra}
\label{types:Hydrae}

\vspace{4pt}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#Hydra}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
Inductive Hydra : Set :=
|  node :  Hydrae -> Hydra
with Hydrae : Set :=
| hnil : Hydrae
| hcons : Hydra -> Hydrae -> Hydrae.
\end{Coqsrc}

%\index{To do}

\begin{todo}
Look for an existing library on trees with nodes of arbitrary arity, in order to replace  this ad-hoc type with something more generic.
\end{todo}

\index{Projects}

\begin{project}[**]

 Another very similar representation could use the \texttt{list} type family instead of the specific 
type \texttt{Hydrae}:

\begin{Coqalt}
Module Alt.

Inductive Hydra: Set :=
  hnode (daughters : list Hydra).

End Alt.
\end{Coqalt}

Using this representation, re-define all the constructions of this chapter.
You will probably have to use patterns described for instance in~\cite{BC04} or the archives of the Coq-club~\cite{Coq}.

\end{project}

\index{Projects}

\begin{project}
The type \texttt{Hydra} above describes hydras as \emph{plane trees}, i.e., as drawn on a sheet of paper or computer screen. Thus, hydras are \emph{oriented},
and it is appropriate to consider a \emph{leftmost} or \emph{rightmost} head of
the beast. It could be interesting to consider another representation, in which
every non-leaf node has a \emph{multi-set} -- not an ordered list -- of daughters.
\end{project}

\subsubsection{Abbreviations}

We provide several notations for ``patterns'' which occur often in our developments.

\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#head}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
(** heads *)
Notation head := (node hnil).

(** nodes  with 1, 2 or 3 daughters *)
Notation hyd1 h := (node (hcons h hnil)).
Notation hyd2 h h' := (node (hcons h (hcons h' hnil))).
Notation hyd3 h h' h'' := 
                   (node (hcons h (hcons h' (hcons h'' hnil)))).
\end{Coqsrc}

For instance, the hydra \texttt{Hy}  of Figure~\vref{fig:Hy} is defined in \emph{Gallina} as follows:

\vspace{4mm}
\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Examples.html\#Hy}{Hydra.Hydra\_Examples}}

\begin{Coqsrc}
Example Hy := hyd3 head
                   (hyd2
                      (hyd1 
                         (hyd2 head head))
                      head) 
                   head.
\end{Coqsrc}

Hydras quite frequently contain  multiple  copies of the same pattern. The following functions
will help us to describe and reason about replications in hydra battles.

\vspace{4pt}

\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Definitions.html\#hcons_mult}{Hydra.Hydra\_Definitions}}

\begin{Coqsrc}
Fixpoint hcons_mult (h:Hydra)(n:nat)(s:Hydrae):Hydrae :=
  match n with 
  | O => s
  | S p => hcons h (hcons_mult h p s)
  end.

(** hydra with n copies of the same daughter *)

Definition hyd_mult h n :=
  node (hcons_mult h n hnil).
\end{Coqsrc}

\vspace{4mm}

For instance, the hydra $Hy''$ of Fig~\vref{fig:Hy3}  can be defined in \coq{} as follows:

\vspace{4pt}

\emph{From Module~\href{../src/html/hydras.Hydra.Hydra_Examples.html}{Hydra.Hydra\_Examples}}

\begin{Coqsrc}
Example Hy'' := 
     hyd2 head
          (hyd2 (hyd_mult (hyd1 head) 5)
                head).
\end{Coqsrc}

\subsubsection{Recursive functions on type \texttt{Hydra}}
\label{sec:orgheadline41}
\label{sec:hsize-def}
